\documentclass[a4paper,twoside,12pt]{article}
\usepackage[left=30mm,right=30mm,top=32mm,bottom=36mm]{geometry}

\usepackage[utf8]{inputenc} % UTF-8
\usepackage[T1]{fontenc} % 8-bit fonts

\usepackage{lmodern} % Modern font

\usepackage{xcolor} % Colors
\usepackage{hyperref} % URLs
\usepackage{graphicx} % Graphics

\usepackage[parfill]{parskip} % Line skip paragraphs

\usepackage{xifthen}
\usepackage{bytefield}

\usepackage{fancyhdr}

\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

% Section styling
\renewcommand{\thepart}{\arabic{part}}
\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\thesubsection}{\hspace{-1em}}

\renewcommand{\familydefault}{\sfdefault} % Sans-seriff font

% command definitions
\input{defines/bytefield}
\input{defines/instructions}

\fancyhf[HLE,HRO]{\leftmark}
\fancyhf[HRE]{CARP}
\fancyhf[HLO]{ISA}

\begin{document}

\pagenumbering{gobble}

\begin{titlepage}
\center

~

\vspace{2cm}

\huge
Cellular Automata Research Platform\\
Instruction Set Architecture

\vfill

\large
Revision 1.0

\large
2015-06-14

\vspace{2cm}

~

\end{titlepage}

\cleardoublepage

\pagestyle{fancy}
\pagenumbering{roman}

\tableofcontents

\cleardoublepage
\pagenumbering{arabic}

\section{Introduction}

This document is a complete specification of the instruction set for the Cellular Automata Research Platform.
It documents all effects and possible side effects of every instruction.

Unless otherwise stated, an instruction completes in one cycle.
However, keep in mind that multi-word instructions require multiple cycles to send over PCI Express.

When a bit vector is broken into multiple words, the least significant part is always listed first.

\clearpage
\subsection{Instructions}

Each instruction is 256 bits and consists of a 5-bit operation code, a 3-bit length field and up to 248 bits of parameters.

The operation code specifies what kind of instruction it is, and how the parameters should be parsed.

The length field is used to improve communication speed by only transmitting the necessary parts of an instruction;
It is zero-extended back to 256 bits by the fetch module.
The field directly specifies the number of words after the first that are sent.

The parameters are of different types and lengths for each instruction.
Please see the individual instruction pages.

\subsubsection*{Instruction Format}

\wordTI{Parameters (low)}{Length}{OpCode}
\wordQ{Parameters (higher)}
$$...$$

\clearpage
\subsection{Rules}

Rules consists of conditions for each cell in the neighborhood and a result that will be applied to the cell if the conditions match.

Each condition contains a type, a state and a bit for each that marks if it should be checked.
The result format is identical except for that the check bits are exchanged with change bits that mark which parts of the cell should change if all conditions match.

In the formats below, [type bits] is assumed to be 5 and [states bits] 1 for the purpose of having everything nicely align to bytes.

\subsubsection*{Rule Format}

\wordSSSS{Condition X–}{Condition X+}{Condition Self}{Result}
\wordSSSS{Condition Z–}{Condition Z+}{Condition Y–}{Condition Y+}

\subsubsection*{Condition Format}

\rulecondition{Type}{Check T}{State}{Check S}

\subsubsection*{Result Format}

\rulecondition{Type}{Change T}{State}{Change S}

\notes

For a rule to be counted as a hit, all conditions must match and at least one change bit must be set.

Conditions for Z are ignored when [matrix depth] is 1.

\clearpage
\subsection{LUTs}

The indexing for the look-up tables is (Z–,Z+,Y–,Y+,X–,X+,Self).
For each of these indexes, the next cell state is specified.
The least significant index is written first (to the right).

In the format below, [state bits] is assumed to be 1 since it is the only value currently supported.
This allows the entries for (Y–,Y+,X–,X+,Self) to fit exactly within one word.

\subsubsection*{LUT Format}

\wordQ{(Y–,Y+,X–,X+,Self) when (Z–,Z+) is 00}
\wordQ{(Y–,Y+,X–,X+,Self) when (Z–,Z+) is 01}
\wordQ{(Y–,Y+,X–,X+,Self) when (Z–,Z+) is 10}
\wordQ{(Y–,Y+,X–,X+,Self) when (Z–,Z+) is 11}

\notes

The Z parts are ignored when [matrix depth] is 1.

\cleardoublepage
\section{General Instructions}

This section covers instructions that are not used directly or do not fit into any of the other categories.

\input{instructions/nop}

\input{instructions/read-information}
\input{instructions/read-fitness}

\input{instructions/swap-cell-storage}
\input{instructions/reset-buffers}

\cleardoublepage
\section{Development Instructions}

This section covers all instructions affecting the development module.
This includes writing rules, setting active rules, running development and reading data for which rules have triggered.

\input{instructions/read-rule-vectors}
\input{instructions/read-rule-numbers}

\input{instructions/write-rule}
\input{instructions/set-rules-active}

\input{instructions/develop}

\cleardoublepage
\section{Cell Storage Instructions}

This section covers all instructions for writing and reading states and types to/from the cell storage.

\input{instructions/read-state-one}
\input{instructions/read-state-all}
\input{instructions/read-type-one}
\input{instructions/read-type-all}

\input{instructions/write-state-one}
\input{instructions/write-state-row}
\input{instructions/write-type-one}
\input{instructions/write-type-row}

\input{instructions/fill-cells}

\cleardoublepage
\section{Cellular Automaton Instructions}

This section covers all instructions affecting the Cellular Automaton.
This includes writing look-up tables, configuring the CA, running the CA, and reading back the new states.

\input{instructions/write-lut}

\input{instructions/config}
\input{instructions/readback}

\input{instructions/step}

\cleardoublepage
\section{Control Flow Instructions}

This section covers all instructions that are related to the program memory.
This includes those for storing, starting and exiting programs, in addition to control flow within the programs.

\input{instructions/break}

\input{instructions/store}
\input{instructions/end}

\input{instructions/jump}
\input{instructions/jump-equal}
\input{instructions/counter-increment}
\input{instructions/counter-reset}

\end{document}
